FlowFPX: Nimble Tools for Debugging Floating-Point Exceptions

### 2023-08-22 paper.tex
Fix library name formatting conventions, resolve feedback from Xinyi

Every arithmetic operation requires a similar overloading.
Thus, the implementation of \FT{} uses a metaprogramming technique
adapted from \texttt{Sherlogs.jl}~\cite{kMilanklSherlogsJl2021}
to abstract over the common patterns.
For every binary operation, the library creates an overloading similar
to the one for \code{+} above.
Unary operations and others work analogously.
This approach saved thousands of lines of code.
The implementation weighs in at 218 lines and defines 645 overloaded function;
assuming 5 lines of code per function, a handwritten version would require over 3,000 lines.

@@

\texttt{ShallowWaters.jl} simulates the flow of water over a
seabed~\cite{klowerNumberFormatsError2020,klowerPositsAlternativeFloats2019}.
The library has dozens of parameters that a scientist can experiment with.
One notable parameter is the Courant-Friedrichs-Lewy (CFL) number, which roughly describes the size of the time step to take in running the simulation.
A small CFL number makes the simulation run slowly, but accurately; a large number speeds it up but loses precision
because the system does not get enough time to propagate information.

@@

Running \FT{} on the high-CFL simulation reveals where the simulation drifted
from numbers to \NaN{} exceptions.
The first step in applying \FT{} is to convert relevant floats to tracked
floats, e.g., \code{Float32} to \code{TrackedFloat32}.
For shallow waters, this step is easy because the simulation is parameterized by
a \fp{} type for internal use.
Swapping in a tracked type is enough:

@@

While the logs for shallow waters contain useful information,
there is an overwhelming amount of it.
There are over ten thousand lines in the gen file alone.
Converting these logs to a stack graph gives a quick overview
of the most common paths to exceptional values.

\Cref{fig:sw_nan_cstg} presents the stack graph for \NaN{} gen events
in shallow waters with the high CFL number.
Reading bottom-up, every \NaN{} came from calls to the `\code{-}' and `\code{+}' operations.
Calls to `\code{+}' account for most of the \NaN{}s.
These \NaN{}s arose in two different contexts: a small number (30)
occurred within a momentum calculation, while the rest (162)
occurred within a continuity step.
Moving to the top of the graph, it shows that the function \code{run\_model} drove
the entire simulation.
With this overview of the program, a promising next step is to guard against \NaN{}s
in the momentum and continuity functions.

\begin{figure}[t]
  \centering
  \includegraphics[width=0.96\columnwidth]{fig/sw_nan_cstg_clean.pdf}
  \caption{Stack graph for \NaN{} gens in shallow waters}
  \label{fig:sw_nan_cstg}
\end{figure}

@@

Graph diffing works well for \CSTG{}s; it shows how flows in the program evolved
from one stage to another.
\Cref{fig:cstg_diff_demo} illustrates one diff in the context of \NaN{} gens for
shallow waters.
It compares the first 10\% of gen events to the latter 90\% of gens.
The positive numbers and green lines indicate flows that are new in the latter part.
The negative numbers and red lines show flows that disappeared in the latter part.

Another one of CSTGs abilities is \emph{graph diffing}.
We took the \FT{} logs from running ShallowWaters and split the logs at the 10\% mark.
We used CSTG to diff the flow graphs between the first 10\% of the logs and the remaining 90\%.
The diff can be seen in \cref{fig:cstg_diff_demo}.
A domain expert might use these clues to find where an instability started in
the first part of the program.
In this case, the function \texttt{momentum\_u!} showed up only in the
beginning of the logs; it may be an effective point to check for \Nan{}s.

@@

With the targeted fuzzing abilities of \FT{}, we set focus on the
\texttt{NBodySimulator} package from the SciML
team.\footnote{\url{https://github.com/SciML/NBodySimulator.jl}}
Fuzzing discovered zero bugs and zero exceptional events.
In fact, fuzzing injected zero exceptional values whatsoever, even with maximum odds.
The reason for these null results, it turns out, is that the nbody
simulation merely creates a problem for the \code{OrdinaryDiffEq}
solver.\footnote{\url{https://github.com/SciML/OrdinaryDiffEq.jl}}
All the \fp{} operations were in another library.

@@

Oceananigans~\cite{OceananigansJOSS} is simulation package for incompressible
fluid dynamics that can generate code for Nvidia GPUs.
For example, the following program (from the project readme) simulates turbulance:

@@

In addition to the cost incurred by intercepting \fp{} operations, gathering stack traces is expensive.
We observe a 10x slowdown on shallow waters with logging disabled.
We recommend that users of \FT{} make use of the \texttt{maxLogs} and \texttt{exclude\_stacktrace} configuration options to limit the number and kind of logs gathered, and thereby reduce the number of calls to \texttt{stacktrace()}.
Stack traces are essential to decide where to inject a fault when fuzzing,
but we defer them as late as possible to maximize performance.

>>>>

Every arithmetic operation requires a similar overloading.
Thus, the implementation of \FT{} uses a metaprogramming technique
adapted from \texttt{Sherlogs}~\cite{kMilanklSherlogsJl2021}
to abstract over the common patterns.
For every binary operation, the library creates an overloading similar
to the one for \code{+} above.
Unary operations and others work analogously.
This approach saved thousands of lines of code.
The implementation weighs in at 218 lines and defines 645 overloaded function;
assuming 5 lines of code per function, a handwritten version would require over 3,000 lines.

@@

\texttt{ShallowWaters} simulates the flow of water over a
seabed~\cite{klowerNumberFormatsError2020,klowerPositsAlternativeFloats2019}.
The library has dozens of parameters that a scientist can experiment with.
One notable parameter is the Courant-Friedrichs-Lewy (CFL) number, which roughly describes the size of the time step to take in running the simulation.
A small CFL number makes the simulation run slowly, but accurately; a large number speeds it up but loses precision
because the system does not get enough time to propagate information.

@@

Running \FT{} on the high-CFL simulation reveals where the simulation drifted
from numbers to \NaN{} exceptions.
The first step in applying \FT{} is to convert relevant floats to tracked
floats, e.g., \code{Float32} to \code{TrackedFloat32}.
For \texttt{ShallowWaters}, this step is easy because the simulation is parameterized by
a \fp{} type for internal use.
Swapping in a tracked type is enough:

@@

While the logs for \texttt{ShallowWaters} contain useful information,
there is an overwhelming amount of it.
There are over ten thousand lines in the gen file alone.
Converting these logs to a stack graph gives a quick overview
of the most common paths to exceptional values.

\Cref{fig:sw_nan_cstg} presents the stack graph for \NaN{} gen events
in \texttt{ShallowWaters} with the high CFL number.
Reading bottom-up, every \NaN{} came from calls to the `\code{-}' and `\code{+}' operations.
Calls to `\code{+}' account for most of the \NaN{}s.
These \NaN{}s arose in two different contexts: a small number (30)
occurred within a momentum calculation, while the rest (162)
occurred within a continuity step.
Moving to the top of the graph, it shows that the function \code{run\_model} drove
the entire simulation.
With this overview of the program, a promising next step is to guard against \NaN{}s
in the momentum and continuity functions.

\begin{figure}[t]
  \centering
  \includegraphics[width=0.96\columnwidth]{fig/sw_nan_cstg_clean.pdf}
  \caption{Stack graph for \NaN{} gens in \texttt{ShallowWaters}}
  \label{fig:sw_nan_cstg}
\end{figure}

@@

Graph diffing works well for \CSTG{}s; it shows how flows in the program evolved
from one stage to another.
\Cref{fig:cstg_diff_demo} illustrates one diff in the context of \NaN{} gens for
\texttt{ShallowWaters}:
it compares the first 10\% of gen events to the latter 90\% of gens.
The positive numbers and green lines indicate flows that are new in the latter part,
and the negative numbers and red lines show flows that disappeared in the latter part.
A domain expert might use these clues to find where an instability started in
the first part of the program.
In this case, the function \texttt{momentum\_u!} showed up only in the
beginning of the logs---it may be an effective point to check for \Nan{}s.

@@

With the targeted fuzzing abilities of \FT{}, we set focus on the
\texttt{NBodySimulator} package from the SciML
team.\footnote{\url{https://github.com/SciML/NBodySimulator.jl}}
We only wanted to test the robustness of the \texttt{NBodySimulator} package, so we configured the fuzzer to inject \NaN{}s \emph{only} when inside of a function of that library.
This way, we could avoid injecting in the standard library or other dependencies of \texttt{NBodySimulator} and concentrate on improving this one library.

However, Fuzzing discovered zero bugs and zero exceptional events.
In fact, fuzzing injected \emph{zero} exceptional values whatsoever, even with adjusting the odds to \emph{force} an injection at the first available opportunity.
The reason for this lack of \NaN{} injection, it turns out, is that the nbody
simulation merely creates a problem for the \code{OrdinaryDiffEq}
solver.\footnote{\url{https://github.com/SciML/OrdinaryDiffEq.jl}}
All the \fp{} operations were in another library.

@@

\texttt{Oceananigans}~\cite{OceananigansJOSS} is simulation package for incompressible
fluid dynamics that can generate code for Nvidia GPUs.
For example, the following program (from the project readme) simulates turbulance:

@@

In addition to the cost incurred by intercepting \fp{} operations, gathering stack traces is expensive.
We observe a 10x slowdown on \texttt{ShallowWaters} with logging disabled.
We recommend that users of \FT{} make use of the \texttt{maxLogs} and \texttt{exclude\_stacktrace} configuration options to limit the number and kind of logs gathered, and thereby reduce the number of calls to \texttt{stacktrace()}.
Stack traces are essential to decide where to inject a fault when fuzzing,
but we defer them as late as possible to maximize performance.
